Determining the existance and direction of causality in the face of measurement error

### 2016-05-07 docs/manuscript.Rmd
updated manuscript

With rapid acceleration in technological advances that enable us to measure the human phenome, we are presented with the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data. For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene all depend on understanding the causal nature of any putative observational associations. Despite the ubiquity of the question, cross-sectional study designs are notoriously susceptible to pitfalls making causal inference, with simple regression based techniques unable to distinguish a true causal association from reverse causation or confounding.

An established tool that can resolve this problem is randomisation. Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the association by randomly splitting the sample into two groups, applying the exposure to one group and leaving the other as a control. The association between the exposure and the outcome in this setting is then guaranteed to be a robust estimate of the causal relationship. In practical terms randomisation is often impossible to implement in such a controlled manner, but the principles can be employed using existing cross sectional data through the use of genetic instruments.

Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling (SEM) [@Aten2008]. A Mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable. If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the independent variable through the mediator.

This simplistic scenario, however, is seldom borne out in reality. Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator. An important tool that has shown promise in resolving this problem is randomisation,

 The principle behind mediation analysis can be explained as follows. Supposing

 Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods, which have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].

While mediation based approaches are simple and convenient to implement, it is important to note where they may lead to unreliable results. One such mechanism is measurement error [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013], which has been noted to be an explicit issue with CIT style approaches [@Wang2015; @Lagani2015]. Measurement (or observational) error is the difference between a measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect sensitivity, so for example methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Sensitivity is also an issue, for example if the unit of measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error will of course arise in more low-tech data too, for example when measuring body mass index (BMI) one is typically interested in using this as a proxy for obesity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically have both bias and imprecision compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].

An alternative statistical method that gained prominence as a solution to the problem of measurement error is Mendelian randomisation (MR) [@DaveySmith2004; @DaveySmithHemani2014]. Here an instrumental variable (typically a SNP) that has a robost, causal association with the exposure is used to stand for the exposure in an association test with the outcome. If there is no measurement error in the SNP then the causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure. Indeed this approach was in part pioneered (in the form of instrumental variable (IV) analysis) as a correction for measurement error [@Ashenfelter1994], but it also has the useful property that it guards against unmeasured confounders driving the association between exposure and outcome; and if the biological effect of the SNP is understood then it can also aid against reverse causality driving the association.

Here, however, arises a potential issue: often the biological effect of a SNP is not known. If there exists a putative association between two variables, with a SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. For example genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. To illustrate a trivial example - the same FTO variant that is famously associated with BMI is also associated with C-reactive protein (CRP). It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would infer that CRP causes BMI. Another situation is where the causal direction between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?

Mediation based methods, for example CIT and other related approaches typically attempt to use the data to resolve this issue using the following logic: if the SNP-exposure association is stronger than the SNP-outcome association then the causal effect must be in the direction of exposure to outcome. The same logic can indeed be applied to MR, but often when applied the sensitivity of these approaches to measurement error is not considered.

When being objective one must assume that measurement error is ubiquitous, and any measured variable is only an imperfect proxy of the biological quantity that the researcher intended to obtain. Here we show using theory and simulations how measurement error can lead to unreliable causal inference in the mediation-based CIT method, and we propose an extension to MR that, although does not eliminate the problem of measurement error, provides a framework in which to ascertain the causal direction when the biology of the instrument is not fully understood.

@@

The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016]. The *cit.cp* function was used to obtain an omnibus p-value of causality between an exposure $x$ and an outcome $y$ using a mediator $g$ by making inferences based on a number of conditions that we examine in the results. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and for incorrect model of $y$ causing $x$. For some significance threshold $\alpha$, causal direction was inferred based on the following scenarios:

1. If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then the correct model is accepted
2. If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then the incorrect model is accepted
3. If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then the an alternative causal model is accepted
4. If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no inference is made

For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value.

@@

Supposing there are two correlated variables, $x$ and $y$, with a variant $g$ associated with both. Assuming no horizontal pleiotropy (i.e. $g$ only has an effect on one of the variables through the other variable) it is desirable to know which of the variables it has a direct influence on. This can be achieved by testing for a difference in the correlations $\rho_{g, x}$ and $\rho_{g, y}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980]. It is calculated as

@@

and a p-value is generated from the $Z$ value to indicate confidence of difference in correlations $\rho_{gx}$ and $\rho_{gy}$. The test for a causal association is then obtained by performing two sample MR in the direction that is predicted to be the correct model based on the Z score test. For the purposes of compiling simulation results, the test is deemed to have obtained a robust association if the Z score test has a p-value $< 0.05$ and the two sample MR analysis has a p-value $< 0.05$, though in practice we do not advocate using an arbitrary $p$ value threshold for denoting significance in making causal inference.

Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting, where the Steiger test of two independent correlations is applied instead of the case detailed here where correlations share one variable in common. An advantage of using the Steiger test in the two sample context is that it can compare correlations in independent samples where sample sizes are different.

@@

$$
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \beta_g g + \epsilon_g \\
x_O & = \beta_{mx} x + \epsilon_{mx} \\
y & = \beta_x x + \epsilon_x \\
y_O & = \beta_{my} y + \epsilon_{my} \\
\end{aligned}
$$

All $\beta$ values were set to 1. Values of $\epsilon_*$ were generated such that

$$
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
var(\epsilon_{my}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
$$

giving a total of 432 combinations of parameters. Each of these sets of variables was performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simualted variables. Similar patterns of results were obtained for different values of $cor(g, x)$.

@@

Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_O=f(x)$. For example, we can define $f(x) = \alpha_m + \beta_m x + \epsilon_m$, where $\alpha_m$ and $\beta_m$ influence the bias in the measurement of $x$, and $\epsilon_m$ represents the imprecision in the measurement of $x$. Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that

@@

where $\beta_g$ is the effect of the SNP on the exposure, and $\epsilon_g$ is the residual value of $x$; and the outcome is partially explained by the exposure

@@

where $\beta_x$ is the true effect of the exposure on the outcome. In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $g$ is associated with $x|y$; and 4) $g$ is independent of $y|x$. The 4th condition is necessary for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$. When measurement error is introduced we can show using basic covariance properties (Appendix 1) that

$$
\begin{aligned}
cov(g, y - \hat{y}) & = cov(g, y) - cov(g, \hat{y})  \\
                    & = \beta_g \beta_x var(g) - D\beta_g \beta_x var(g)
\end{aligned}
$$

@@

$$
D = \frac{\beta^2_m var(x)} {\beta^2_m var(x) + var(\epsilon_m)}
$$

Thus an observational study will find $cov(g, y - \hat{y}) = 0$ when the true model is causal only when $D = 1$. Therefore, if there is any measurement error that incurs imprecision (i.e. $var(\epsilon_m) \neq 0$) then there will remain an association between $g$ and $y|x$, which is in violation of the the 4th condition of the CIT. Note that measurement bias alone is insufficient to lead to a violation of the test statistic assumptions.

We performed simulations to verify that this problem does arise using the CIT method. `r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_O}=1$) the CIT is reliable in identifying the correct causal direction. However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction. Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the evidence of the incorrect inference.

@@

$$
\beta_{MR} = \frac{\beta_y}{\beta_g} = \frac{cov(y, g)}{cov(x_O, g)} = \frac{cov(\beta_x x + \epsilon_x, g)}{cov(x + \epsilon_x, g)}
$$

which reduces to

$$
\lim_{n \to \infty} \hat{\beta}_{MR} = \frac{\beta_x cov(x + \epsilon_x)}{cov(x + \epsilon_x)} = \beta_x
$$

thus we obtain an estimate of the effect of $x_O$ on $y$ that is unbiased by measurement error in $x$.

However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$. But it can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is important to note that under certain conditions when measurement error is present the Steiger test could make erroneous inference about causal direction because the value of $d$ is not restricted to being greater than 0. However, using the MR approach there are two potential advantages over the CIT. First, we can conduct a formal test for the direction of causality. Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome. `r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.

>>>>

The last decade has seem a rapid acceleration in technological advances that enable us to measure the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits. This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data. For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene expression all depend on understanding the causal nature of any putative observational associations. Despite the ubiquity of the question, commonly used cross-sectional study designs are notoriously susceptible to pitfalls in making causal inference, with simple regression-based techniques unable to distinguish a true causal association from reverse causation or confounding.

An established statistical design that can resolve this problem is randomisation. Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This is the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant cross sectional data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels. Two statistical approaches to exploiting the properties of genetic instruments are commonly used: mediation and Mendelian randomisation (MR)  [@DaveySmith2004; @DaveySmithHemani2014].

Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling (SEM) [@Aten2008]. A mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable. If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the dependent variable through the mediator.

This simplistic scenario, however, is seldom borne out in reality. Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator. A proposed solution to this is to use a genetic instrument as the independent variable. Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods. They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].

MR employs the same data under a completely different model. Here an instrumental variable (typically a SNP) that has a robost, direct association with the exposure is used as a surrogate for the exposure in an association test with the outcome. Here the problems of confounding or reverse causation are avoided by testing for association between the outcome and the SNP. The causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.

By utilising genetic instruments in different ways, mediation and MR models have properties that confer some advantages and some disadvantages. The main advantage of mediation is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables.

Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure. This requirement can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. For example, the same FTO variant that is associated with body mass index (BMI) is also associated with C-reactive protein (CRP). It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI. Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?

MR does however have some important advantages over mediation based approaches. First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size. Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR relax this assumption, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes. Third, and though this is often underappreciated in fields outside classical epidemiology it is of crucial importance, MR is substantially more robust to there being measurement error in the exposure variable. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both basic mediation [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence. We argue that this is a serious limitation.

Measurement (or observational) error is the difference between the measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, so for example methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error will of course arise in more low-tech data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both bias and imprecision compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].

Given the near-ubiquitous presence of measurement error in biological data it is vital to understand its impact on the tools we use for causal inference. Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method. We also present an extension to MR that allows researchers to ascertain the causal direction even when the biology of the instrument is not fully understood. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially.

@@

The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016]. The methodology of the CIT is as follows. Assume an exposure $x$ is instrumented by a SNP $g$, and the exposure $x$ is causally related to an outcome $y$. Thus,

$$
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_g + \beta_g g + \epsilon_g \\
y & = \alpha_x + \beta_x x + \epsilon_x \\
\end{aligned}
$$

The following tests are then performed:

1. $H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0$; *the SNP associates with the exposure*
2. $H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0$; *the SNP associates with the outcome*
3. $H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0$; *the exposure associates with the outcome*
4. $H_0: cov(g, y - \hat{y}) \neq 0; cov(g, y - \hat{y}) = 0$; *the SNP is independent of the outcome when the outcome is adjusted for the exposure*

where $y - \hat{y} = y - \hat{alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for the $x$, where $x$ is assumed to mediate the association between the SNP and the outcome. If all four tests reject the null hypothesis then it is inferred that $x$ is causally related to $y$. The CIT measures the strength of causality by generating an omnibus p-value, $p_{CIT}$, which is simply the largest (least extreme) p-value of the four tests, the intuition being that causal inference is only as strong as the weakest link in the chain of tests.

In these analyses the *cit.cp* function was used to obtain an omnibus p-value. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and for incorrect model of $y$ causing $x$. For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:

- If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then the correct model is accepted
- If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then the incorrect model is accepted
- If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then an alternative causal model is accepted
- If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no inference is made

For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value, though we should stress that for real analyses it is not good practice to use p-values for making causal inference, nor is it reliable to depend on arbitrary significance thresholds.

@@

Two stage least squares (2SLS) is a commonly used technique for performing MR when the exposure, outcome and instrument data are all available in the same sample. Assuming the following model

$$
y = \alpha_{MR} + \beta_{MR} \hat{x} + \epsilon_{MR}
$$

where $\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g$, a causal relationship is inferred if the null hypothesis that $\beta_{MR} = 0$ is rejected. A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$.

The direction of the causal association was inferred as follows. Assuming no horizontal pleiotropy (i.e. $g$ only has an effect on one of the variables through the other variable) it is desirable to know which of the variables, $x$ or $y$, it has a direct influence on. This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (Appendix 2). This test can be formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980]. It is calculated as

@@

and a p-value is generated from the $Z$ value to indicate the probability of obtaining the observed difference in correlations $\rho_{gx}$ and $\rho_{gy}$ under the null hypothesis that both correlations are identical. The existence of causality and its direction is inferred based on the following scenarios:

- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z > 0$ then a causal association for the correct model is accepted
- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z < 0$ then a causal association for the incorrect model is accepted
- Otherwise no inference is made

For the purposes of compiling simulation results, we use an arbitrary $\alpha = 0.05$ value.

Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting, where the Steiger test of two independent correlations is applied instead of the case detailed here where correlations share one variable in common. An advantage of using the Steiger test in the two sample context is that it can compare correlations in independent samples where sample sizes are different. The method and R function are provided in Appendix 3.

@@

$$
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_x + \beta_x x + \epsilon_x \\
y_O & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
$$

where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, and $*_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $*_{my}$ are parameters for non-differential measurement error in the outcome $y$.

All $\alpha$ values were set to 0, and $\beta$ values set to 1. Values of $\epsilon_*$ were generated such that

$$
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
$$

giving a total of 432 combinations of parameters. Simulations using each of these sets of variables were performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simulated variables. Similar patterns of results were obtained for different values of $cor(g, x)$.

@@

Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_O = f(x)$. For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$, and $\epsilon_{mx}$ represents the imprecision in the measurement of $x$. Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that

@@

where $\beta_g$ is the effect of the SNP on the exposure, and $\epsilon_g$ is the normally distributed residual value of $x$; and the outcome is partially explained by the exposure

@@

where $\beta_x$ is the true effect of the exposure on the outcome. In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$. The 4th condition is necessary for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$. When measurement error is introduced we can show using basic covariance properties (Appendix 1) that

$$
\begin{aligned}
cov(g, y - \hat{y}) & = cov(g, y_O) - cov(g, \hat{y}_O)  \\
                    & = \beta_{my} \beta_g \beta_x var(g) - D \beta_{my} \beta_g \beta_x var(g)
\end{aligned}
$$

@@

$$
D = \frac{\beta^2_{mx} var(x)} {\beta^2_{mx} var(x) + var(\epsilon_{mx})}
$$

Thus an observational study will find $cov(g, y_O - \hat{y_O}) = 0$ when the true model is causal only when $D = 1$. Therefore, if there is any measurement error that incurs imprecision (i.e. $var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_O | x$, which is in violation of the the 4th condition of the CIT. Note that linear transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions.

We performed simulations to verify that this problem does arise using the CIT method. `r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_O}=1$) the CIT is reliable in identifying the correct causal direction. However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction. Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.

@@

$$
\begin{aligned}
\beta_{MR} = \frac{\beta_{y_O}}{\beta_g} & = \frac{cov(y_O, g)}{cov(x_O, g)} \\
                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
\end{aligned}
$$

thus we obtain an estimate of the effect of $x_O$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision. Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate, but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.

However, if we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result. This is because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$. But it can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is important to note that under certain conditions when measurement error is present the Steiger test could make erroneous inference about causal direction because the value of $d$ is not restricted to being greater than 0. However, using the MR approach there are two potential advantages over the CIT. First, we can conduct a formal test for the direction of causality. Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome. `r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.
